Run Life expectancy and decomposition calculations

### 2017-01-20 Code/2_Run_LE_and_Decomp_Calcs.Rmd

```{r load_libraries, echo=F, warning=F, message=F}
library(tidyverse)
library(plotly)
options(tibble.width = Inf) #display all columns of tibbles
```

@@

```{r load_data}
dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", header = T)
dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", header = T)
```

@@

```{r calculate_life_expectancy}
non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))

@@

    list.life.tables1[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
                                                 num.ages.in.group = "Age_bin_size",
                                                 death.counts = "total_deaths1",
                                                 population.counts = "Population")

    list.life.tables5[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
                                                  num.ages.in.group = "Age_bin_size",
                                                  death.counts = "total_deaths5",
                                                  population.counts = "Population")

    list.life.tables9[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, StateYearRaceSex == stratum.id) %>% arrange(Age_factor),
                                                  num.ages.in.group = "Age_bin_size",
                                                  death.counts = "total_deaths9",
                                                  population.counts = "Population")
    list.index <- list.index + 1
    print(list.index)
  }
)
```

@@

les <- tibble(le1, le5, le9, sid)
names(les) = c("life.expectancy.birth.1", "life.expectancy.birth.5", "life.expectancy.birth.9", "StateYearRaceSex")

dat.aggregated <- merge(dat.aggregated, les, by = "StateYearRaceSex", all.x =T)

@@

```{r densityplot_life_expectancy}
ggplot(dat.aggregated %>% dplyr::filter(Age_factor == "<1 year"), aes(x = life.expectancy.birth.1)) + 
  geom_density(aes(fill="Impute 1"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
  facet_grid(Sex_factor ~ Race_factor)
```

@@

Black <- dat.aggregated %>% filter(Race_factor == "Black" & Age_minbin == 0) %>% 
                            select(State_factor, Year_actual, Sex_factor, Pop_across_age, 
                                   life.expectancy.birth.1, life.expectancy.birth.5, 
                                   life.expectancy.birth.9) %>%
                            rename(Pop_black = Pop_across_age,  
                                   le.birth.black1 = life.expectancy.birth.1, 
                                   le.birth.black5 = life.expectancy.birth.5, 
                                   le.birth.black9 = life.expectancy.birth.9)

White <- dat.aggregated %>% filter(Race_factor == "White" & Age_minbin == 0) %>% 
                            select(State_factor, Year_actual, Sex_factor, Pop_across_age, 
                                   life.expectancy.birth.1, life.expectancy.birth.5, 
                                   life.expectancy.birth.9) %>%
                            rename(Pop_white = Pop_across_age,  
                                   le.birth.white1 = life.expectancy.birth.1,
                                   le.birth.white5 = life.expectancy.birth.5,
                                   le.birth.white9 = life.expectancy.birth.9)

BlackWhite <- merge(Black, White, by = c("State_factor", "Year_actual", "Sex_factor"))

@@

#first, create list of stratum.ids for each race-sex combination
for (rs in unique(levels(dat.clean$RaceSex))) {
  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
         dat.clean %>% 
           ungroup() %>%
           filter(RaceSex == rs) %>%
           select(StateYearRaceSex, StateYearSex, Sex_factor, State_factor, Year_actual) %>%
           group_by(StateYearRaceSex, StateYearSex) %>%
           filter(row_number() == 1)
  )

@@

#then, merge the black-white lists
Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
                    by = "StateYearSex", suffixes = c("White", "Black"))
head(Female.ids)

Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
                  by = "StateYearSex", suffixes = c("White", "Black"))
head(Male.ids)

@@

system.time(
  for(i in 1:dim(paired.ids)[1]) {
    white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
    black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

    list.age.decomp.tables1[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
    list.age.decomp.tables5[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
    list.age.decomp.tables9[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]

    if(length(white.index) != 0 & length(black.index) != 0) {
      list.age.decomp.tables1[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "Age_factor")

      list.age.decomp.tables5[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "Age_factor")

      list.age.decomp.tables9[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "Age_factor")
    } else {
      list.age.decomp.tables1[[i]]$decomp.table <- list.age.decomp.tables5[[i]]$decomp.table <- list.age.decomp.tables9[[i]]$decomp.table <-"Couldn't compare life tables"
    }
  }
)

@@

system.time(
  for(i in 1:dim(paired.ids)[1]) {
    white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
    black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

    list.cod.decomp.tables1[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
    list.cod.decomp.tables5[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]
    list.cod.decomp.tables9[[i]]$StateYearSex <- paired.ids[i, "StateYearSex"]

    if(length(white.index) != 0 & length(black.index) != 0) {
      ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
        ungroup() %>%
        select(Age_factor, COD_factor, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
        rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)

      ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
        ungroup() %>%
        select(Age_factor, COD_factor, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
        rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)

      ct.both <- merge(ct.white, ct.black, by = c("Age_factor", "COD_factor")) 
      ct.both <- ct.both %>% arrange(COD_factor, Age_factor)

@@

        list.cod.decomp.tables1[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]$life.table),
                                                              life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables1[[i]]$decomp.table),
                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white1, prop.black1), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black1", 
                                                              prop2.colname.cod.table = "prop.white1")

        list.cod.decomp.tables5[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table),
                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white5, prop.black5), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black5", 
                                                              prop2.colname.cod.table = "prop.white5")

        list.cod.decomp.tables9[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table),
                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
                                                              cod.table = ct.both %>% select(Age_factor, COD_factor, prop.white9, prop.black9), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black9", 
                                                              prop2.colname.cod.table = "prop.white9")

@@

```{r pretty_cod}
ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))

ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))
```

>>>>

```{r load_libraries, echo=F, warning=F, message=F}
library(tidyverse)
options(tibble.width = Inf) #display all columns of tibbles
```

@@

```{r load_data}
dat.aggregated <- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", header = T)
dat.clean<- read.csv("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", header = T)
```

@@

```{r calculate_life_expectancy}
non.zero.stratums <- setdiff(levels(dat.aggregated$state.year.race.sex), 
                             unique(dat.aggregated$state.year.race.sex[dat.aggregated$population == 0]))

@@

    list.life.tables1[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, state.year.race.sex == stratum.id) %>% arrange(age),
                                                 num.ages.in.group = "age_bin_size",
                                                 death.counts = "total_deaths1",
                                                 population.counts = "population")

    list.life.tables5[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, state.year.race.sex == stratum.id) %>% arrange(age),
                                                  num.ages.in.group = "age_bin_size",
                                                  death.counts = "total_deaths5",
                                                  population.counts = "population")

    list.life.tables9[[list.index]]$life.table <- life.table(data = subset(dat.aggregated, state.year.race.sex == stratum.id) %>% arrange(age),
                                                  num.ages.in.group = "age_bin_size",
                                                  death.counts = "total_deaths9",
                                                  population.counts = "population")
    list.index <- list.index + 1
    print(list.index)
  }
)
```

@@

les <- tibble(le1, le5, le9, sid)
names(les) = c("life.expectancy.birth.1", "life.expectancy.birth.5", "life.expectancy.birth.9", "state.year.race.sex")

dat.aggregated <- merge(dat.aggregated, les, by = "state.year.race.sex", all.x =T)

@@

```{r densityplot_life_expectancy}
ggplot(dat.aggregated %>% dplyr::filter(age == "<1 year"), aes(x = life.expectancy.birth.1)) + 
  geom_density(aes(fill="Impute 1"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
  facet_grid(sex ~ race)
```

@@

Black <- dat.aggregated %>% filter(race == "Black" & age_minbin == 0) %>% 
                            select(state, year.n, sex, pop_across_age, 
                                   life.expectancy.birth.1, life.expectancy.birth.5, 
                                   life.expectancy.birth.9) %>%
                            rename(pop_black = pop_across_age,  
                                   le.birth.black1 = life.expectancy.birth.1, 
                                   le.birth.black5 = life.expectancy.birth.5, 
                                   le.birth.black9 = life.expectancy.birth.9)

White <- dat.aggregated %>% filter(race == "White" & age_minbin == 0) %>% 
                            select(state, year.n, sex, pop_across_age, 
                                   life.expectancy.birth.1, life.expectancy.birth.5, 
                                   life.expectancy.birth.9) %>%
                            rename(pop_white = pop_across_age,  
                                   le.birth.white1 = life.expectancy.birth.1,
                                   le.birth.white5 = life.expectancy.birth.5,
                                   le.birth.white9 = life.expectancy.birth.9)

BlackWhite <- merge(Black, White, by = c("state", "year.n", "sex"))

@@

#first, create list of stratum.ids for each race-sex combination
for (rs in unique(levels(dat.clean$race.sex))) {
  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
         dat.clean %>% 
           ungroup() %>%
           filter(race.sex == rs) %>%
           select(state.year.race.sex, state.year.sex, sex, state, year.n) %>%
           group_by(state.year.race.sex, state.year.sex) %>%
           filter(row_number() == 1)
  )

@@

#then, merge the black-white lists
Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
                    by = "state.year.sex", suffixes = c(".white", ".black"))
head(Female.ids)

Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
                  by = "state.year.sex", suffixes = c(".white", ".black"))
head(Male.ids)

@@

system.time(
  for(i in 1:dim(paired.ids)[1]) {
    white.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.white"])
    black.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.black"])

    list.age.decomp.tables1[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
    list.age.decomp.tables5[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
    list.age.decomp.tables9[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]

    if(length(white.index) != 0 & length(black.index) != 0) {
      list.age.decomp.tables1[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "age")

      list.age.decomp.tables5[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "age")

      list.age.decomp.tables9[[i]]$decomp.table <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]$life.table), 
                                                                 name.lt1 = "Black", 
                                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]$life.table), 
                                                                 name.lt2 = "White", age.groups = "age")
    } else {
      list.age.decomp.tables1[[i]]$decomp.table <- list.age.decomp.tables5[[i]]$decomp.table <- list.age.decomp.tables9[[i]]$decomp.table <-"Couldn't compare life tables"
    }
  }
)

@@

system.time(
  for(i in 1:dim(paired.ids)[1]) {
    white.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.white"])
    black.index <- which(non.zero.stratums == paired.ids[i, "state.year.race.sex.black"])

    list.cod.decomp.tables1[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
    list.cod.decomp.tables5[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]
    list.cod.decomp.tables9[[i]]$state.year.sex <- paired.ids[i, "state.year.sex"]

    if(length(white.index) != 0 & length(black.index) != 0) {
      ct.white <- subset(dat.clean, state.year.race.sex == paired.ids[i, "state.year.race.sex.white"]) %>% 
        ungroup() %>%
        select(age, COD, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
        rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)

      ct.black <- subset(dat.clean, state.year.race.sex == paired.ids[i, "state.year.race.sex.black"]) %>% 
        ungroup() %>%
        select(age, COD, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
        rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)

      ct.both <- merge(ct.white, ct.black, by = c("age", "COD")) 
      ct.both <- ct.both %>% arrange(COD, age)

@@

        list.cod.decomp.tables1[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables1[[black.index]]$life.table),
                                                              life.table2 = as.data.frame(list.life.tables1[[white.index]]$life.table), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables1[[i]]$decomp.table),
                                                              cod.table = ct.both %>% select(age, COD, prop.white1, prop.black1), 
                                                              age.colname.cod.table = "age", 
                                                              COD.colname.cod.table = "COD", 
                                                              prop1.colname.cod.table = "prop.black1", 
                                                              prop2.colname.cod.table = "prop.white1")

        list.cod.decomp.tables5[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table),
                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
                                                              cod.table = ct.both %>% select(age, COD, prop.white5, prop.black5), 
                                                              age.colname.cod.table = "age", 
                                                              COD.colname.cod.table = "COD", 
                                                              prop1.colname.cod.table = "prop.black5", 
                                                              prop2.colname.cod.table = "prop.white5")

        list.cod.decomp.tables9[[i]]$decomp.table <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]$life.table),
                                                              life.table2 = as.data.frame(list.life.tables5[[white.index]]$life.table), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables5[[i]]$decomp.table),
                                                              cod.table = ct.both %>% select(age, COD, prop.white9, prop.black9), 
                                                              age.colname.cod.table = "Age_factor", 
                                                              COD.colname.cod.table = "COD_factor", 
                                                              prop1.colname.cod.table = "prop.black9", 
                                                              prop2.colname.cod.table = "prop.white9")

@@

```{r pretty_cod}
ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
  geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
  xlab("Contribution to life expectancy gap") +
  ylab("Age group\n") + theme_minimal() +
  geom_vline(xintercept = 0)

ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
  geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
  xlab("Contribution to life expectancy gap") +
  ylab("Age group\n") + theme_minimal() +
  geom_vline(xintercept = 0)
```
